Detailed model description

### 2021-11-30 doc/appendix-model.Rmd
Start model appendix

### Matern parameterization

@@

>>>>

```{r setup, echo=FALSE, cache=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = FALSE,
  error = FALSE,
  message = FALSE,
  warning = FALSE
)
```

\newcommand{\s}{\boldsymbol{s}}

# Notation conventions

This appendix uses the following notation conventions:

* Bold lowercase for vectors

* Bold subscript $\s$ since x, y

* Bold uppercase for matrices

* $\phi$ for all 'dispersion' parameters for consistency with code

* Family titles link to TMB or sdmTMB source code

* Attempt to link math symbols to argument options in model and code symbols

```{r, results='asis'}
symbols <- tibble::tribble(
  ~Symbol, ~Description,
  "$\\phi$", "A dispersion parameter for a distribution",
  "$\\mu$", "Mean",
  "$y$", "Observed response data",
  "$f$", "Link function",
  "$\\omega_{\\s}$", "Spatial random field at point $\\s$",
  "$\\zeta_{\\s, t}$", "Spatially varying coefficient random field at point $\\s$ and time $t$",
  "$\\epsilon_{\\s, t}$", "Spatiotemporal random field",
  "$\\boldsymbol{\\Sigma}_\\omega$", "Spatial random field covariance matrix",
  "$\\boldsymbol{\\Sigma}_\\epsilon$", "Spatiotemporal random field covariance matrix",
  "$\\boldsymbol{\\Sigma}_\\zeta$", "Spatially varying coefficient random field covariance matrix",
  "$\\boldsymbol{\\beta}$", "Parameter vector",
  "$\\boldsymbol{X}$", "Model predictor matrix",
  "$\\alpha_g$", "IID random intercept deviations",
  "$\\sigma_\\alpha^2$", "IID random intercept variance",
  "$\\delta_{\\s,t}$", "AR1 or random walk spatiotemporal deviations",
  "$\\rho$", "Correlation between random fields in subsequent time steps"

)
knitr::kable(symbols, caption = "Symbol notation and descriptions", escape = FALSE, format = "latex", booktabs = TRUE)
```

```{r}
subscripts <- tibble::tribble(
  ~Symbol, ~Description,
  "$\\s$", "Index for space; a vector of x and y coordinates",
  "$t$", "Index for time",
  "$g$", "Group"
)
knitr::kable(subscripts, caption = "Subscript notation", escape = FALSE, format = "latex", booktabs = TRUE)
```

# Basic model structure

$$
\begin{aligned}
  \mathbb{E}(y_{\s,t}) &= \mu_{\s,t},\\
  \mu_{\s,t} &= f^{-1}
  \left( \boldsymbol{X} \boldsymbol{\beta} + \omega_s + \epsilon_{\s,t} \right),\\
  \boldsymbol{\omega} &\sim \operatorname{MVNormal} \left( \boldsymbol{0}, \boldsymbol{\Sigma}_\omega \right),\\
  \boldsymbol{\epsilon}_t &\sim \operatorname{MVNormal} \left( \boldsymbol{0}, \boldsymbol{\Sigma}_{\epsilon} \right).
\end{aligned}
$$
where $f$ is a link function (e.g., log or logit) and $f^{-1}$ is its inverse link.

* $\boldsymbol{X} \boldsymbol{\beta}$ is defined by the `formula` argument and
  represents the fixed effect model matrix and a corresponding vector of
  coefficients

* $\omega_s$ are included if `spatial = 'on'`

* $\epsilon_{\s,t}$ are included by default if there are multiple time elements (`time` argument is not `NULL`) but can be set to IID (`IID`; default), AR1 (`AR1`), random walk (`RW`), or off (`'off`) via the `spatiotemporal` argument

# Random or multi-level intercepts

$$
\begin{aligned}
  \mu_{\s,t} &= f^{-1} \left( \ldots + \alpha_{g} + \ldots \right),\\
  \alpha_g &\sim \operatorname{Normal} \left(0, \sigma_\alpha^2 \right),\\
\end{aligned}
$$

where $\alpha_g$ is an example optional "random" intercept---an intercept
centered on zero that varies by level $g$ and is constrained by
$\sigma_\alpha$. This is defined by the `formula` argument via the `(1 | g)`
syntax as in lme4 or glmmTMB.

# Time-varying regression parameters

$$
\begin{aligned}
  \mu_{\s,t} &= f^{-1} \left( \ldots + \gamma_{t} x_{\s,t} + \ldots \right),\\
  \gamma_{t=1} &\sim \operatorname{Uniform} \left(-\infty, \infty \right),\\
  \gamma_{t>1} &\sim \operatorname{Normal} \left(\gamma_{t-1}, \sigma^2_{\gamma} \right),
\end{aligned}
$$

where $\gamma_t$ is an example optional time-varying regression parameter and
$x_{\s,t}$ is the corresponding predictor value. This is defined via the
`time_varying` argument. E.g., `~ 0 + x`. The first value is given an implicit
uniform prior. I.e., the same variable should not appear in the fixed effect
formula since the initial value is estimated as part of the time-varying formula.

# Spatial regression parameters

$$
\begin{aligned}
  \mu_{\s,t} &= f^{-1} \left( \ldots + \zeta_s x_t + \ldots \right),\\
  \boldsymbol{\zeta} &\sim \operatorname{MVNormal} \left( \boldsymbol{0}, \boldsymbol{\Sigma}_\zeta \right),
\end{aligned}
$$

where $\zeta_s$ is a regression parameter that varies spatially and $x_t$ is a predictor that is constant for a given time $t$. Currently, $x_t$ must be the time variable itself. If you are interested in using a different predictor, contact the developers.

# AR1 spatiotemporal random fields

Dropping the optional $\omega_s$ for simplicity:

$$
\begin{aligned}
\mu_{\s,t} &= f^{-1} \left( \boldsymbol{X} \boldsymbol{\beta} + \delta_{\s,t} \right),\\
\boldsymbol{\delta}_{t=1} &\sim \operatorname{MVNormal} (\boldsymbol{0}, \boldsymbol{\Sigma}_{\epsilon}),\\
\boldsymbol{\delta}_{t>1} &= \rho \boldsymbol{\delta}_{t-1} + \sqrt{1 - \rho^2} \boldsymbol{\epsilon}_t,  \:
\boldsymbol{\epsilon}_t \sim \operatorname{MVNormal} \left(\boldsymbol{0}, \boldsymbol{\Sigma}_{\epsilon} \right),
\end{aligned}
$$

where $\rho$ is the correlation between subsequent spatiotemporal random
fields. The $\rho \boldsymbol{\delta}_{t-1} + \sqrt{1 - \rho^2}$ scales the spatiotemporal variance by the correlation such that it represents the steady-state marginal variance.

# Random walk spatiotemporal random fields

Dropping the optional $\omega_s$ for simplicity:

$$
\begin{aligned}
\mu_{\s,t} &= f^{-1} \left( \boldsymbol{X} \boldsymbol{\beta} + \delta_{\s,t} \right),\\
\boldsymbol{\delta}_{t=1} &\sim \operatorname{MVNormal} (\boldsymbol{0}, \boldsymbol{\Sigma}_{\epsilon}),\\
\boldsymbol{\delta}_{t>1} &= \boldsymbol{\delta}_{t-1} +  \boldsymbol{\epsilon}_t,  \:
\boldsymbol{\epsilon}_t \sim \operatorname{MVNormal} \left(\boldsymbol{0}, \boldsymbol{\Sigma}_{\epsilon} \right),
\end{aligned}
$$

# Offset terms

Offset terms can be included with the reserved word `offset` in the formula.
E.g., `y ~ x + offset`.

These are included in the linear predictor as

$$
\begin{aligned}
  \mu_{\s,t} &= f^{-1} \left( \ldots + O_{\s,t} + \ldots \right),
\end{aligned}
$$
where $O_{\s,t}$ is an offset term---a **log transformed** variable without
a coefficient (assuming a log link).

# Threshold models

## Linear breakpoint threshold models

TODO

These models can be fit by including `+ breakpt(x)` in the model formula, where
`x` is a covariate.

## Logistic threshold models

The form is

$$
s(x)=\tau + \psi\ { \left[ 1+{ e }^{ -\ln\ \left(19\right) \cdot \left( x-s50 \right)
     / \left(s95 - s50 \right) } \right] }^{-1},
$$
where $\psi$ is a scaling parameter (controlling the height of the y-axis for
the response, and is unconstrained), $\tau$ is an intercept, $s50$ is
a parameter controlling the point at which the function reaches 50% of the
maximum ($\psi$), and $s95$ is a parameter controlling the point at which the
function reaches 95%. The parameter $s50$ is unconstrained, and $s95$ is
constrained to be larger than $s50$.

These models can be fit by including `+ logistic(x)` in the model formula, where `x` is a covariate.

# Observation model families

## [Binomial](https://kaskr.github.io/adcomp/group__R__style__distribution.html#gaee11f805f02bc1febc6d7bf0487671be)

Internally parameterized as the [robust
version](https://kaskr.github.io/adcomp/group__R__style__distribution.html#gaecb5a18095a320b42e2d20c4b120f5f5).

## [Beta](https://kaskr.github.io/adcomp/group__R__style__distribution.html#ga5324c83759d5211c7c2fbbad37fa8e59)

$$
\operatorname{Beta} \left(\mu \phi, 1 - \mu \phi \right)
$$

where $\phi$ is variance.

$$
\operatorname{Binomial} \left( N, \mu \right)
$$

$N = 1$ (i.e., 'size' currently fixed) and $\mu$ is probability

## [Gamma](https://kaskr.github.io/adcomp/group__R__style__distribution.html#gab0e2205710a698ad6a0ed39e0652c9a3)

As shape, scale:

$$
\operatorname{Gamma} \left( \phi, \frac{\mu}{\phi}  \right)
$$

where $\phi$ represents the shape and $\frac{\mu}{\phi}$ represents the scale.

## [Gaussian](https://kaskr.github.io/adcomp/dnorm_8hpp.html)

$$
\operatorname{Normal} \left( \mu, \phi \right)
$$
where $\phi$ is the standard deviation (following Stan convention of SD not variance).

## [Lognormal](https://github.com/pbs-assess/sdmTMB/blob/28a93496a93b12e5a2f73f703bc96afbf14d86c1/src/sdmTMB.cpp#L22-L30)

$$
\operatorname{Lognormal} \left( \log \mu - \frac{\phi^2}{2}, \phi \right)
$$

## [Negative Binomial](https://kaskr.github.io/adcomp/group__R__style__distribution.html#ga76266c19046e04b651fce93aa0810351)

Internally parameterized as the [robust version](https://kaskr.github.io/adcomp/group__R__style__distribution.html#gaa23e3ede4669d941b0b54314ed42a75c)

$$
\operatorname{NB2} \left( \mu, \phi \right)
$$

Variance scales quadratically with mean $\mathrm{Var}[y] = \mu + \mu^2 / \phi$.

## [Poisson](https://kaskr.github.io/adcomp/group__R__style__distribution.html#gaa1ed15503e1441a381102a8c4c9baaf1)

$$
\operatorname{Poisson} \left( \mu \right)
$$

## [Student-t](https://github.com/pbs-assess/sdmTMB/blob/28a93496a93b12e5a2f73f703bc96afbf14d86c1/src/sdmTMB.cpp#L11-L20)

$$
\operatorname{Student-t} \left( \mu, \phi, \nu \right)
$$

where $\nu$, the degrees of freedom, is currently fixed at 3.

## [Tweedie](https://kaskr.github.io/adcomp/group__R__style__distribution.html#ga262f3c2d1cf36f322a62d902a608aae0)

[Source code](https://kaskr.github.io/adcomp/tweedie_8cpp_source.html) as in [cplm](https://CRAN.R-project.org/package=cplm) package. <https://link.springer.com/article/10.1007/s11222-012-9343-7>

$$
\operatorname{Tweedie} \left(\mu, p, \phi \right), \: 1 < p < 2
$$

## Matern parameterization

@@

# References
